Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-17 article.tex
wording with Gilles

In this paper, we describe the winning method\footnote{Code available at
\url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The rest of the
paper is structured as follows: Section \ref{sec:filter} describes the signal
processing methods applied on fluorescent calcium time series; Section
\ref{sec:inference} then presents the proposed approach and its theoretical
properties; Section \ref{sec:results} provides an empirical analysis and
comparison with other network inference methods, while Section
\ref{sec:conclusion} finally discusses our work and provides further research
directions.

\section{Signal filtering} \label{sec:filter}

@@

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to use a low-pass filter for
filtering out high frequency noise, using one of the following designs,
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
x^t_i &\leftarrow x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
x^t_i &\leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

Considering the short delay of communication between neurons and the slow decay
of fluorescence signal, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align}
% an asymmetrical discrete derivative filter
x^{t,d}_{i} &\leftarrow x^{t}_i - x_{t-1}^i. \label{eq:high-pass-filter}
\end{align}

@@

At this step, the processed signal only contains clean spikes. However, when a
large of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 x^{t,w}_i  &\leftarrow (x^{t,d}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t,d}_j}}.
\end{align}

\textcolor{red}{G: Let's drop that? or reformulate...}
\textcolor{red}{PCA?} Finally since there is a high correlation between neurons,
we filter some more noise through a principal component analysis. From the $|V|$
eigen vectors, we kept $80\%$ of the eigen vectors which have the largest eigen
value. In the context of the challenge, this assumption was deduced from ground
truth of provided networks.

@@

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $X = \left\{X_1, \ldots, X_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{X}$, and by
$X^{-i,j}$ the set of all variables except $X_i$ and $X_j$.
Two random variables $X_i$ and $X_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $X_i \perp X_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{X}_i, \mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$X_i \not\perp X_j | X^{-i,j}$ between all pairs $i$ and $j$ of fluorescence
signals, in other words the inferred network will be
$\hat{A}: \hat{a}_{i,j} = \mathbb{1}(X_i \not\perp X_j | X^{-i,j}), \forall i, j \in V$.

@@

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}
\]
is equal to zero \citep{koller2009probabilistic}. The partial correlation gives
a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{X_i, X_j | X^{-i,j}}, \forall i, j \in V$ between all
pairs of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
Similarly, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

@@

>>>>

In this paper, we present a simplified -- yet nearly as good -- version of
the winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The rest of the paper is structured as
follows: Section \ref{sec:filter} describes the signal processing methods
applied on fluorescent calcium time series; Section \ref{sec:inference} then
presents the proposed approach and its theoretical properties; Section
\ref{sec:results} provides an empirical analysis and comparison with other
network inference methods, while Section \ref{sec:conclusion} finally
discusses our work and provides further research directions. Additionnaly, Appendix
\ref{app:optimized} further describes our actual winning method, giving slightly
better results than the method presented in the paper but highly tuned for the
challenge.

\section{Signal processing} \label{sec:filter}

@@

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to smoothen the signal, using a low-pass filter for
filtering out high frequency noise, applying one of the following designs and
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
x^t_i &\leftarrow x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
x^t_i &\leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

Considering the short delay of communication between neurons and the slow
decay of fluorescence, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align} % an asymmetrical discrete derivative filter 
x^{t,d}_{i} &\leftarrow x^{t}_i - x_{t-1}^i. \label{eq:high-pass-filter} 
\end{align}

@@

At this step, the processed signal only contains clean spikes. However, when a
large part of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 x^{t,w}_i  &\leftarrow (x^{t,d}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t,d}_j}}.
\end{align}

@@

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $X = \left\{X_1, \ldots, X_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{X}$, and by
$X^{-i,j}$ the set of all variables except $X_i$ and $X_j$.
Two random variables $X_i$ and $X_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $X_i \perp X_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{X}_i, \mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$X_i \not\perp X_j | X^{-i,j}$ between all pairs $i$ and $j$ of fluorescence
signals.

@@

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}
\]
is equal to zero \citep{koller2009probabilistic}. The partial correlation gives
an estimate of the degree of association between all
pairs of neuron $i$ and $j$. 
Similarly, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

@@

\textcolor{red}{Say that we use PCA to derive the precision matrix + ref!.}

Finally since there is a high correlation between neurons,
we filter some more noise through a principal component analysis. From the $|V|$
eigen vectors, we kept $80\%$ of the eigen vectors which have the largest eigen
value. In the context of the challenge, this assumption was deduced from ground
truth of provided networks.
